---
title: "Explanation of `diego_analysis.R`: Cardiovascular Disease Logistic Regression Analysis"
author: "STAT 4355 Project Analysis"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    number_sections: true
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Purpose of the File

The file `diego_analysis.R` performs a logistic regression analysis on a
cleaned cardiovascular disease dataset. The main response variable is
$$cardio_i =
\begin{cases}
1, & \text{if patient } i \text{ has cardiovascular disease},\\
0, & \text{if patient } i \text{ does not have cardiovascular disease}.
\end{cases}$$

The goal of the script is to identify factors associated with
cardiovascular disease risk and to compare several logistic regression
models. The script also evaluates predictive performance, checks
residual diagnostics, performs threshold selection, explores interaction
terms, and creates plots suitable for presentation.

# Packages and Data Loading

The script begins by installing and loading several R packages:

```{r, eval=FALSE}
install.packages("ggplot2", type = "binary")
install.packages("DHARMa", type = "binary")
install.packages("MASS", type = "binary")
install.packages("olsrr", type = "binary")
install.packages("glmnet")
install.packages("pROC")

library(car)
library(ggplot2)
library(DHARMa)
library(MASS)
library(olsrr)
library(glmnet)
library(pROC)
```

The important packages are:

-   `ggplot2`: creates plots.

-   `DHARMa`: creates simulated residual diagnostics for generalized
    linear models.

-   `glmnet`: fits LASSO logistic regression models.

-   `pROC`: creates ROC curves and calculates AUC.

-   `car`: can be used for variance inflation factors, although the VIF
    code is commented out.

The file then reads in the cleaned dataset:

```{r, eval=FALSE}
readLines("filtered_cardio.csv", n = 5)
t <- read.csv2("C:\\Users\\dxe230000\\Desktop\\filtered_cardio.csv")
names(t)
```

The command `readLines()` checks the first few lines of the file. The
command `read.csv2()` loads the dataset. The function `read.csv2()` is
used for files that are commonly separated by semicolons rather than
commas. The object `t` stores the dataset.

# Commented-Out Preprocessing

Several preprocessing steps are included but commented out. These
include filtering blood pressure values, removing missing observations,
removing unrealistic BMI values, converting height and weight to
imperial units, converting age to years, and creating dummy variables.

For example, the commented-out code includes:

```{r, eval=FALSE}
# t$height_in <- t$height / 2.54 
# t$weight_lbs <- t$weight * 2.20462
# t$age_years <- t$age /365
```

This suggests that the file `filtered_cardio.csv` already contains the
cleaned and transformed variables needed for modeling, such as
`age_years`, `height_imperial`, `weight_imperial`, and dummy variables
for cholesterol, glucose, and gender.

# The First Logistic Regression Model: `cardiomodel`

The first main model is:

```{r, eval=FALSE}
cardiomodel <- glm(cardio ~ age_years + height_imperial + weight_imperial +
                     ap_hi + ap_lo + gender_dv +
                     cholesterol_2_dv + cholesterol_3_dv +
                     gluc_2_dv + gluc_3_dv +
                     smoke + alco + active,
                   data = t, family = binomial)
summary(cardiomodel)
```

Because `family = binomial` is used, this is a logistic regression
model. The model estimates the probability that a patient has
cardiovascular disease: $$p_i = P(cardio_i = 1).$$

The logistic regression model has the form
$$\log\left(\frac{p_i}{1-p_i}\right)
= \beta_0 + \beta_1 age_i + \beta_2 height_i + \beta_3 weight_i + \beta_4 ap\_hi_i + \cdots + \beta_k active_i.$$

The left-hand side is the log-odds of cardiovascular disease. If we
define $$\eta_i = \beta_0 + \beta_1x_{i1} + \cdots + \beta_kx_{ik},$$
then the predicted probability is
$$\hat{p}_i = \frac{1}{1+e^{-\eta_i}}.$$

This first model includes age, height, weight, systolic blood pressure,
diastolic blood pressure, gender, cholesterol dummy variables, glucose
dummy variables, smoking, alcohol use, and activity.

# The Reduced Main-Effects Model: `cardiomodel2`

The second model is a reduced model:

```{r, eval=FALSE}
cardiomodel2 <- glm(cardio ~ age_years + height_imperial + weight_imperial +
                      ap_hi + ap_lo +
                      cholesterol_2_dv + cholesterol_3_dv +
                      gluc_3_dv +
                      smoke + alco + active,
                    data = t, family = binomial)
summary(cardiomodel2)
```

This model removes `gender_dv` and `gluc_2_dv`. It keeps only a smaller
set of predictors that were considered more important for comparison.

The purpose of this model is to create a simpler and more interpretable
logistic regression model. It still models $$P(cardio=1),$$ but with
fewer predictors.

# Comparing `cardiomodel` and `cardiomodel2`

The script compares the full model and the reduced model using AIC and a
likelihood ratio test:

```{r, eval=FALSE}
AIC(cardiomodel, cardiomodel2)
anova(cardiomodel2, cardiomodel, test = "Chisq")
drop1(cardiomodel2, test = "Chisq")
```

## AIC

AIC is the Akaike Information Criterion: $$AIC = 2k - 2\log(L),$$ where
$k$ is the number of estimated model parameters and $L$ is the model
likelihood. Lower AIC indicates a better balance between model fit and
model simplicity.

## Likelihood Ratio Test

The command

```{r, eval=FALSE}
anova(cardiomodel2, cardiomodel, test = "Chisq")
```

compares the reduced model to the full model. The test statistic is
$$G^2 = 2\{\log(L_{full}) - \log(L_{reduced})\}.$$ Under the null
hypothesis, this approximately follows a chi-square distribution:
$$G^2 \sim \chi^2_{df}.$$ The null hypothesis is that the extra
variables in the full model do not significantly improve model fit.

## Single-Term Drop Tests

The command

```{r, eval=FALSE}
drop1(cardiomodel2, test = "Chisq")
```

tests whether each term in `cardiomodel2` can be removed one at a time.
A small p-value means removing that term would significantly worsen the
model.

# DHARMa Residual Diagnostics for `cardiomodel2`

The script then creates simulated residual diagnostics:

```{r, eval=FALSE}
sim_res_model_cm2 <- simulateResiduals(cardiomodel2)
testoutliers_cm2 <- testOutliers(simulationOutput = sim_res_model_cm2, type = "bootstrap")

plotQQunif(sim_res_model_cm2)
plotResiduals(sim_res_model_cm2, smoothScatter = FALSE)

testDispersion(sim_res_model_cm2)
testUniformity(sim_res_model_cm2)
```

For logistic regression, ordinary residuals can be difficult to
interpret because the response is binary. DHARMa simulates response
values from the fitted model and compares the observed data to simulated
data. If the model fits well, the DHARMa residuals should be
approximately uniformly distributed between 0 and 1.

The diagnostics check:

-   Whether the residuals are uniformly distributed.

-   Whether there are too many outliers.

-   Whether there is overdispersion or underdispersion.

-   Whether residual patterns appear across predicted values.

# Train-Test Prediction Using `cardiomodel2`

The script then performs a 70/30 train-test split:

```{r, eval=FALSE}
set.seed(123)
n <- nrow(t)
train_idx <- sample(seq_len(n), size = round(0.7 * n))

train <- t[train_idx, ]
test  <- t[-train_idx, ]
```

The model is trained on 70% of the observations and tested on the
remaining 30%:

```{r, eval=FALSE}
cardiomodel2_train <- glm(
  cardio ~ age_years + height_imperial + weight_imperial +
    ap_hi + ap_lo +
    cholesterol_2_dv + cholesterol_3_dv +
    gluc_3_dv +
    smoke + alco + active,
  data = train,
  family = binomial
)
```

Predicted probabilities are computed on the test set:

```{r, eval=FALSE}
pred_prob <- predict(cardiomodel2_train, newdata = test, type = "response")
```

The output `pred_prob` contains values between 0 and 1:
$$\hat{p}_i = P(cardio_i = 1 \mid x_i).$$

Using a cutoff of 0.5, the script classifies patients:

```{r, eval=FALSE}
pred_class <- ifelse(pred_prob >= 0.5, 1, 0)
cm <- table(Predicted = pred_class, Actual = test$cardio)
```

# Classification Metrics

From the confusion matrix, the script extracts:

-   $TP$: true positives.

-   $TN$: true negatives.

-   $FP$: false positives.

-   $FN$: false negatives.

The metrics are:

$$Accuracy = \frac{TP+TN}{TP+TN+FP+FN},$$

$$Sensitivity = \frac{TP}{TP+FN},$$

$$Specificity = \frac{TN}{TN+FP},$$

$$Precision = \frac{TP}{TP+FP}.$$

Sensitivity is also called recall or the true positive rate. Specificity
is the true negative rate. Precision is the proportion of predicted
disease cases that were actually disease cases.

# Density Plot of Predicted Probabilities

The script creates a density plot comparing predicted probabilities for
the two actual classes:

```{r, eval=FALSE}
plot_data <- data.frame(
  pred_prob = pred_prob,
  cardio = factor(test$cardio, levels = c(0, 1), labels = c("No Disease", "Disease"))
)

ggplot(plot_data, aes(x = pred_prob, fill = cardio, color = cardio)) +
  geom_density(alpha = 0.3) +
  labs(
    title = "Density of Predicted Cardiovascular Disease Probabilities",
    x = "Predicted Probability",
    y = "Density",
    fill = "Actual Class",
    color = "Actual Class"
  ) +
  theme_minimal()
```

This graph shows how well the model separates patients with disease from
patients without disease. A better model would show the disease group
shifted toward higher predicted probabilities and the no-disease group
shifted toward lower predicted probabilities.

# Threshold Tuning

The script evaluates thresholds from 0.01 to 0.99:

```{r, eval=FALSE}
thresholds <- seq(0.01, 0.99, by = 0.01)
```

For each threshold, the script recalculates accuracy, sensitivity,
specificity, precision, Youden's $J$, F1 score, and the gap between
sensitivity and precision.

Youden's statistic is $$J = Sensitivity + Specificity - 1.$$

The F1 score is
$$F1 = \frac{2(Precision)(Sensitivity)}{Precision + Sensitivity}.$$

The script identifies three possible thresholds:

-   The threshold that maximizes Youden's $J$.

-   The threshold that maximizes the F1 score.

-   The threshold where sensitivity and precision are closest.

This is useful because the default cutoff 0.5 may not be the best cutoff
for a medical classification problem.

# Attempted Full Interaction Model: `cardiomodel3`

The script attempts a very large interaction model:

```{r, eval=FALSE}
cardiomodel3 <- glm(formula = cardio ~ (1 + age_years + height_imperial + weight_imperial+ ap_hi+ ap_lo) * 
                      (1 + gender_dv) *
                      (1 + smoke_dv) * 
                      (1 + alco_dv) * 
                      (1 + active_dv) * 
                      (1 + cholesterol_2_dv + cholesterol_3_dv) * 
                      (1 + gluc_2_dv + gluc_3_dv), data = t, family = binomial)
```

This formula includes all main effects and many higher-order
interactions. The `*` operator in R expands to main effects and
interactions. For example, $$(a+b)^2$$ in formula notation does not mean
squaring variables; it means including main effects and pairwise
interactions.

This model did not converge. This is expected because the model is
extremely large and contains many high-order interactions. Such a model
can suffer from overfitting, separation, collinearity, and convergence
problems.

# Forward Selection Attempts: `cardiomodel6`, `cardiomodel6b`, and `cardiomodel6c`

The script then tries to add terms to `cardiomodel2` using likelihood
ratio tests.

## Interaction-Block Forward Selection

The first attempt uses:

```{r, eval=FALSE}
model_try <- update(cardiomodel6, as.formula(paste("~ . *", term)))
```

This adds the candidate term and its interactions with all existing
model terms. For each candidate, the script compares the current model
and candidate model using a chi-square likelihood ratio test:

```{r, eval=FALSE}
a <- anova(cardiomodel6, model_try, test = "Chisq")
```

The candidate with the smallest significant p-value is added. This
process repeats until no candidate improves the model at
$\alpha = 0.05$. However, this approach can be slow because each step
may add many interaction terms.

## Main-Effect Forward Selection

The second attempt uses:

```{r, eval=FALSE}
model_try <- update(cardiomodel6b, as.formula(paste("~ . +", term)))
```

This adds only the candidate main effect instead of all interactions.
However, because many of the candidate variables are already included in
`cardiomodel2`, this attempt mostly leaves the model unchanged.

## Selected Candidate Forward Selection

The third attempt tests a smaller set of candidate main effects and
interactions, such as:

```{r, eval=FALSE}
"cholesterol_3_dv:gluc_3_dv"
"age_years:active"
"weight_imperial:active"
"ap_hi:ap_lo"
```

This is a more targeted approach, but the script comments that the
resulting model remains effectively equal to `cardiomodel2`.

## Important Coding Note

In the loops for `cardiomodel6b` and `cardiomodel6c`, the code assigns
the updated model to `cardiomodel6` rather than `cardiomodel6b` or
`cardiomodel6c`:

```{r, eval=FALSE}
cardiomodel6 <- models[[best_term]]
```

If the intention was to update `cardiomodel6b` or `cardiomodel6c`, this
line should be changed accordingly. Otherwise, `summary(cardiomodel6b)`
and `summary(cardiomodel6c)` may still show the unchanged base model.

# LASSO Logistic Regression with Pairwise Interactions

Because the full interaction model is too large, the script uses LASSO
regression:

```{r, eval=FALSE}
x <- model.matrix(
  ~ (age_years + height_imperial + weight_imperial +
       ap_hi + ap_lo +
       gender_dv +
       cholesterol_2_dv + cholesterol_3_dv +
       gluc_2_dv + gluc_3_dv +
       smoke + alco + active)^2,
  data = t
)[, -1]
```

The formula includes all main effects and all pairwise interactions, but
not higher-order interactions. If there are 13 predictors, then the
number of pairwise interactions is
$$\binom{13}{2} = \frac{13(12)}{2}=78.$$ Thus the model matrix contains
roughly $$13 + 78 = 91$$ predictor columns.

LASSO logistic regression solves the penalized optimization problem
$$\min_{\beta}\left[-\log L(\beta) + \lambda \sum_{j=1}^{p}|\beta_j|\right].$$
The penalty term $\lambda \sum |\beta_j|$ shrinks some coefficients
exactly to zero. Terms with zero coefficients are removed from the
model.

The cross-validated LASSO model is fit using:

```{r, eval=FALSE}
cardiolasso_cv <- cv.glmnet(
  x, y,
  family = "binomial",
  alpha = 1,
  nfolds = 10,
  standardize = TRUE
)
```

Here, `alpha = 1` specifies LASSO. The script uses 10-fold
cross-validation to choose $\lambda$. It then uses `lambda.1se`, which
chooses a simpler model within one standard error of the best
cross-validation error:

```{r, eval=FALSE}
coef_1se <- coef(cardiolasso_cv, s = "lambda.1se")
```

# Selected Terms and the Hierarchy Principle

The script extracts nonzero LASSO coefficients:

```{r, eval=FALSE}
selected_terms <- rownames(coef_1se)[as.vector(coef_1se !=0)]
selected_terms <- setdiff(selected_terms, "(Intercept)")
```

If LASSO selects an interaction term, the script also includes its
parent main effects:

```{r, eval=FALSE}
interaction_terms <- selected_terms[grepl(":", selected_terms)]

if (length(interaction_terms) > 0) {
  main_effects_interactions <- unique(unlist(strsplit(interaction_terms, ":")))
  selected_terms <- unique(c(selected_terms, main_effects_interactions))
}
```

This follows the hierarchy principle. For example, if the model includes
$$ap\_hi:cholesterol\_3\_dv,$$ it should also include
$$ap\_hi \quad \text{and} \quad cholesterol\_3\_dv.$$

This makes the interaction model easier to interpret.

# Post-LASSO Logistic Regression: `cardiomodel8`

After LASSO selects terms, the script refits those terms with ordinary
logistic regression:

```{r, eval=FALSE}
lasso_formula <- as.formula(
  paste("cardio ~", paste(selected_terms, collapse = " + "))
)

cardiomodel8 <- glm(lasso_formula, data = t, family = binomial)
```

This is called a post-LASSO model. LASSO is used for variable selection,
and then ordinary logistic regression is used to estimate coefficients
and p-values without the LASSO shrinkage penalty.

# Pruning the Post-LASSO Model

The script then prunes `cardiomodel8` by keeping:

-   significant interaction terms with p-values below 0.05,

-   significant main effects with p-values below 0.05,

-   parent main effects for any kept interaction.

This creates `cardiomodel8_reduced`:

```{r, eval=FALSE}
cardiomodel8_reduced <- glm(
  cardiomodel8_reduced_formula,
  data = t,
  family = binomial
)
```

The final model is compared to the original post-LASSO model using AIC
and a likelihood ratio test:

```{r, eval=FALSE}
AIC(cardiomodel2, cardiomodel8_reduced, cardiomodel8)
anova(cardiomodel8_reduced, cardiomodel8, test = "Chisq")
drop1(cardiomodel8_reduced, test = "Chisq")
```

The goal is to see whether the reduced post-LASSO model keeps most of
the predictive value while remaining simpler than the full post-LASSO
model.

# Residual Diagnostics for the LASSO-Based Models

The script performs DHARMa diagnostics for both `cardiomodel8` and
`cardiomodel8_reduced`:

```{r, eval=FALSE}
sim_res_model_cm8 <- simulateResiduals(cardiomodel8)
plotQQunif(sim_res_model_cm8)
plotResiduals(sim_res_model_cm8, smoothScatter = FALSE)
testDispersion(sim_res_model_cm8)
testUniformity(sim_res_model_cm8)

sim_res_model_cm8_r <- simulateResiduals(cardiomodel8_reduced)
plotQQunif(sim_res_model_cm8_r)
plotResiduals(sim_res_model_cm8_r, smoothScatter = FALSE)
testDispersion(sim_res_model_cm8_r)
testUniformity(sim_res_model_cm8_r)
```

These diagnostics check whether the final logistic models show evidence
of poor fit, outliers, non-uniform residuals, or dispersion problems.

# ROC Curve and AUC

The script evaluates the reduced post-LASSO model using an ROC curve:

```{r, eval=FALSE}
probs <- as.numeric(unlist(predict(cardiomodel8_reduced, type = "response")))
actuals <- as.numeric(unlist(t$cardio))
roc_obj <- roc(actuals, probs)
auc(roc_obj)
plot(roc_obj, col= "lightblue", lwd=3, main = "ROc Curve for cm8_reduced")
```

The ROC curve plots sensitivity against false positive rate across all
possible thresholds. The false positive rate is
$$FPR = \frac{FP}{FP+TN}.$$

The AUC measures how well the model separates patients with
cardiovascular disease from patients without cardiovascular disease. A
common interpretation is:

-   AUC = 0.50: random guessing.

-   AUC around 0.70: acceptable separation.

-   AUC around 0.80 or higher: strong separation.

-   AUC = 1.00: perfect separation.

# Odds Ratio Plot for Presentation

The script creates an odds ratio plot for `cardiomodel2`:

```{r, eval=FALSE}
or_table <- data.frame(
  term = names(coef(cardiomodel2)),
  OR = exp(coef(cardiomodel2)),
  lower = exp(confint(cardiomodel2)[,1]),
  upper = exp(confint(cardiomodel2)[,2])
)
```

For a logistic regression coefficient $\beta_j$, the odds ratio is
$$OR_j = e^{\beta_j}.$$

Interpretation:

-   If $OR_j > 1$, the predictor increases the odds of cardiovascular
    disease.

-   If $OR_j < 1$, the predictor decreases the odds of cardiovascular
    disease.

-   If $OR_j = 1$, the predictor has no association with the odds of
    disease.

The 95% confidence interval is created by exponentiating the coefficient
confidence interval: $$e^{\hat{\beta}_j \pm 1.96SE(\hat{\beta}_j)}.$$

The plot includes a red dashed line at $OR=1$. Terms with confidence
intervals entirely above or below 1 are easier to interpret as
statistically significant risk-increasing or risk-decreasing predictors.

# AIC Comparison Plot

The script creates a bar plot comparing AIC values for:

-   the main-effects model `cardiomodel2`,

-   the full post-LASSO model `cardiomodel8`,

-   the reduced post-LASSO model `cardiomodel8_reduced`.

The model with the lowest AIC has the best balance of goodness of fit
and simplicity among the compared models.

# Important Interactions Identified

The script comments that the reduced post-LASSO model selected several
important interactions, including:

-   `age_years:ap_hi`,

-   `ap_hi:cholesterol_3_dv`,

-   `weight_imperial:ap_hi`,

-   `cholesterol_3_dv:gluc_3_dv`.

For example, the interaction `age_years:ap_hi` means the effect of
systolic blood pressure on cardiovascular disease risk may depend on
age. A model containing this interaction has the form
$$\eta_i = \cdots + \beta_1 age_i + \beta_2 ap\_hi_i + \beta_3(age_i \cdot ap\_hi_i).$$
Then the effect of systolic blood pressure on the log-odds of disease is
$$\frac{\partial \eta_i}{\partial ap\_hi_i} = \beta_2 + \beta_3 age_i.$$
Thus, if $\beta_3\neq 0$, the effect of systolic blood pressure changes
with age.

The interaction `cholesterol_3_dv:gluc_3_dv` means the combined effect
of well-above-normal cholesterol and well-above-normal glucose may
differ from the sum of their separate effects.

# Overall Interpretation of the Script

Overall, `diego_analysis.R` performs a complete logistic regression
workflow for cardiovascular disease risk:

1.  Load the cleaned cardiovascular disease dataset.

2.  Fit a full main-effects logistic regression model.

3.  Fit a smaller, more interpretable main-effects logistic regression
    model.

4.  Compare models using AIC and likelihood ratio tests.

5.  Check residual diagnostics using DHARMa.

6.  Split the data into training and testing sets.

7.  Compute predicted probabilities and classification metrics.

8.  Tune the probability threshold using Youden's statistic, F1 score,
    and sensitivity-precision balance.

9.  Attempt a very large interaction model, which does not converge.

10. Use LASSO logistic regression to search through pairwise
    interactions without making the model explode.

11. Refit LASSO-selected terms using ordinary logistic regression.

12. Prune the post-LASSO model while preserving interaction hierarchy.

13. Evaluate final models using AIC, DHARMa diagnostics, ROC curves, and
    AUC.

14. Create odds ratio and AIC plots for presentation.

# Recommended Summary for a Presentation

A concise explanation of the file is:

> This script models cardiovascular disease status using logistic
> regression. It first fits and evaluates a main-effects model using
> age, body measurements, blood pressure, cholesterol, glucose, smoking,
> alcohol use, and activity. Model quality is assessed using AIC,
> likelihood ratio tests, classification metrics, and DHARMa residual
> diagnostics. The script then explores more complex interaction models.
> Since a full interaction model fails to converge, LASSO logistic
> regression is used to select important pairwise interactions. The
> selected terms are refit in a regular logistic regression model,
> pruned while preserving hierarchy, and evaluated using ROC/AUC and
> diagnostic plots. Odds ratio and AIC plots are produced for final
> presentation.

# Potential Improvements

A few improvements could make the script cleaner and easier to
reproduce:

-   Avoid running `install.packages()` every time the script runs;
    package installation should usually be separate from analysis.

-   Replace the absolute file path with a relative path, such as
    `read.csv2("filtered_cardio.csv")`.

-   Avoid naming the dataset `t`, since `t()` is also a base R transpose
    function.

-   Fix the update line in the `cardiomodel6b` and `cardiomodel6c` loops
    if those models are intended to update themselves.

-   Store model comparison results and classification metrics in clean
    tables for reporting.

-   Use consistent naming for variables and models.
